Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-02-11 Response.tex

\begin{itemize}

% 1
\item {
{\it
\textbf{
The thesis starts with a very nice introduction, which I think is very insightful for the reader who is not familiar with phylogenetic and/or Bayesian inference. 
In spite of this, I believe that the readability of the later chapters would be further improved if the introduction:
}% END: textbf
}% END: it
}% END: item

\begin{itemize}

% 1a
\item {
{\it
\textbf{
Briefly described the evolutionary models, such as GY94, that are later used.
}% END: textbf
}% END: it
}% END: item
\\
I have now added a \emph{Markov models of codon evolution} subsection in the introductory chapter, which discusses the most popular codon model implementations. The first paragraph of this new section is reproduced below (references are not reproduced correctly using the Latex `copy-over' tools we use in this response document):

@@

\begin{quote}
\myeditsvonea
\end{quote}

% 1b
\item {
{\it
\textbf{
Briefly described how to accommodate time-stamped sequences, which is now considered in the analyses in Chapter 2, but not really discussed.
}% END: textbf
}% END: it
}% END: item
\\
%
%PL: TODO: rewrite the final paragraph to reflect the dated-tip likelihood model: Essential, we now consider heights (instead of branch lengths) and introduce a new parameter, the rate of evolution r, which allows rescaling genetic distance into time. The task is now to estimate the heights and r from the sequence data, keeping the ti[ heights proportional to their sampling time, with the constant of proportionality being our estimable r.
%
I added a subsection which describes the methods used in BEAST to accommodate time-stamped sequences as a source for calibrating trees, when such information is available. The first paragraph of this new section is reproduced below:

@@

\begin{quote}
\myeditsvoneb
\end{quote}

% 1c
\item {
{\it
\textbf{
Give a bit more detail on phylogeography, e.g. how do the models look like and what is the connection between phylogeography and epoch models (some connection is described in Chapter 2, but remains a bit vague).
}% END: textbf
}% END: it
}% END: item
\\
I added a figure and additional explanation, which should improve the flow of subsequent chapters. 
These changes start with:

### 2015-02-11 chapters/chapter01.tex
revision editing

Phylogeography is a research direction aimed at connecting evolutionary processes, which happen over time, with the processes of spatial dispersal into a joint spatio-temporal dynamic that can be used to track pathogen spread.
%GB: I don't like the use of 'reciprocal' here ...
% FB: could also use complementary
Historically, phylogeographic inference in molecular epidemiology ignored the correlated character of these two processes. 
The evolutionary relationship represented by the tree was inferred omitting the geographical information, and then the ancestral locations generated by the process of the spatial diffusion were reconstructed, conditioning on the inferred tree topology.
\citet{Lemey2009} first considered the Bayesian reconstruction that allows for joint inference of both a time-measured evolutionary history and the parameters of the spatial and temporal process that shapes the spread and evolution of 
% e.g. 
rapidly evolving viruses.
This approach does not need to fix the tree topology, the branch lengths nor the Markov model parameters, and allows for integration over all possible values, requiring only for a specification of a prior distributions for the parameters that we aim to infer.

\myedit{svonec}{
Central to the Bayesian phylogeographic framework is the hypothesis that sequence evolution is occurring simultaneously with the process of geographical dispersal. 
The observed data, residing at the $N$ tips of phylogeny $\mathbf{F}$, is recorded as both character sequence data $\mathbf{X}=(X_{1},...,X_{N})$ and discretized geographic locations (countries, districts, cities) $\mathbf{Y}=(Y_{1},...,Y_{N})$, as pictured in the upper facette of Figure~\ref{fig:discrete_phylo_illust}.
We assume independent stochastic processes are responsible for generating these data, with ancestral states of $\mathbf{X}$ generated as before by a CTMC characterized by rate matrix $\mathbf{Q}$ and the unobserved ancestral locations $(Y_{N+1},...,Y_{2N-2})$ generated by a separate CTMC characterized by rate matrix $\mathbf{\Phi}$ (see Figure~\ref{fig:discrete_phylo_illust}, lower facette).
The entries of this matrix are the transition rates $\phi_{ij}$ between locations $i$ and $j$.
Unlike before, the rate matrix does not necessarily have to be symmetrical, i.e. there exist some $i,j$ such that $\phi_{ij}\neq\phi_{ji}$.
Additionally, a priori one would expect many of the infinitesimal rates to be zero. 
}

%---DISCRETE PHYLOGEOGRAOHY ILLUSTRATION---%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.4]{discretePhylogeography} 
\caption{
{ \footnotesize 
{\bf Discrete phylogeographical model between $K=4$ states.} 
Data consist of both the character sequences and locations at which they were recorded.
Model allows for non-symmetric rates of exchanged between locations.
}% END: footnotesize
}
\label{fig:discrete_phylo_illust}
\end{figure}

Unlike the molecular data, which is typically dense, the location data is sparse, with only one observation per taxon and many connections not observed at all.
The Bayesian stochastic search variable selection (BSSVS, see also Subsection~\ref{sub:clocks}) procedure is used to allow exchange rates in $\mathbf{\Phi}$ that do not contribute to the diffusion to be shrunken to zero with a certain probability.

The Bayesian framework described in Subsections~\ref{sub:bayesTheorem} and \ref{sub:mcmc} offers the unique ability to integrate different sources of information about the viruses with their genetic data, without conditioning on either one of them.
We use this ability to infer the tree topology, the ancestral spatial locations and the character substitution process giving rise to the aligned molecular sequence data. 
We assume independent stochastic processes are responsible for generating these data.
This allows us to write down the joint posterior distribution:

@@

The likelihood component for both molecular data $\mathbf{X}$ and location data $\mathbf{Y}$ is calculated using recursive tree pruning, as described in Subsection~\ref{sub:likelihood}.
We approximate the joint posterior as written down in Equation~(\ref{eq:posteriorPhylogeography}) using MCMC methods we discussed in Subsection~\ref{sub:mcmc}, which leads to the ancestral reconstruction of the most plausible histories, as depicted in the left facette of the Figure~\ref{fig:discrete_phylo_illust}.

>>>>

Phylogeography is a research direction aimed at connecting evolutionary processes, which happen over time, with the processes of spatial dispersal into a joint spatio-temporal dynamic that can be used to track pathogen spread.
%GB: I don't like the use of 'reciprocal' here ...
% FB: could also use complementary
Historically, phylogeographic inference in molecular epidemiology ignored the correlated nature of these two processes. 
The evolutionary relationships represented by the tree were inferred were first estimated from the sequence data, and then the ancestral locations generated by the process of the spatial diffusion were reconstructed from the spatial data, conditioning on the inferred tree topology.
\citet{Lemey2009} first considered the Bayesian reconstruction that allows for joint inference of both a time-measured evolutionary history and the parameters of the spatial and temporal process that shapes the spread and evolution of 
% e.g. 
rapidly evolving viruses.
This approach does not require fixing the tree topology, the branch lengths nor the Markov model parameters, and allows for integration over all possible values, requiring only for a specification of a prior distributions for the parameters that we aim to infer.

\myedit{svonec}{
Central to the Bayesian phylogeographic framework is the hypothesis that sequence evolution is occurring simultaneously with the process of geographical dispersal. 
The observed data, residing at the $N$ tips of phylogeny $\mathbf{F}$, is recorded as both character sequence data $\mathbf{X}=(X_{1},...,X_{N})$ and discretized geographic locations (countries, districts, cities) $\mathbf{Y}=(Y_{1},...,Y_{N})$, as depicted in the upper facet of Figure~\ref{fig:discrete_phylo_illust}.
We assume independent stochastic processes are responsible for generating these data, with ancestral states of $\mathbf{X}$ generated as before by a CTMC characterized by rate matrix $\mathbf{Q}$ and the unobserved ancestral locations $(Y_{N+1},...,Y_{2N-2})$ generated by a separate CTMC characterized by rate matrix $\mathbf{\Phi}$ (see Figure~\ref{fig:discrete_phylo_illust}, lower facet).
The entries of this matrix are the transition rates $\phi_{ij}$ between locations $i$ and $j$.
Unlike before, the rate matrix does not necessarily have to be symmetrical, i.e. there exist some $i,j$ such that $\phi_{ij}\neq\phi_{ji}$.
Additionally, a priori one would expect many of the infinitesimal rates to be zero. 
}

%---DISCRETE PHYLOGEOGRAOHY ILLUSTRATION---%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.4]{discretePhylogeography} 
\caption{
{ \footnotesize 
{\bf Discrete phylogeographical model for $K=4$ location states.} 
The data consist of both the character sequences and locations at which they were recorded.
The model allows for non-symmetric rates of location exchange in the evolutionary history.
}% END: footnotesize
}
\label{fig:discrete_phylo_illust}
\end{figure}

Unlike the molecular data, which is typically dense, the location data is sparse, with only one observation per taxon and many transitions are likely to remain observed.
For this reason, a Bayesian stochastic search variable selection (BSSVS, see also Subsection~\ref{sub:clocks}) procedure has been proposed in order to allow exchange rates in $\mathbf{\Phi}$ that do not contribute to the phytogeographic diffusion to shrink to zero with a certain probability.

The Bayesian framework described in Subsections~\ref{sub:bayesTheorem} and \ref{sub:mcmc} offers the unique ability to integrate different sources of information about the viruses with their genetic data, without conditioning on either one of them.
We use this ability to infer the tree topology, the ancestral spatial locations and the character substitution process giving rise to the aligned molecular sequence data. 
Since assume independent stochastic processes are responsible for generating these data, we can write down the joint posterior distribution as:

@@

The likelihood component for both molecular data $\mathbf{X}$ and location data $\mathbf{Y}$ is calculated using recursive tree pruning, as described in Subsection~\ref{sub:likelihood}.
We approximate the joint posterior as written down in Equation~(\ref{eq:posteriorPhylogeography}) using MCMC methods we discussed in Subsection~\ref{sub:mcmc}, which leads to the reconstruction of the most plausible histories with ancestral location states, as depicted in the left facet of the Figure~\ref{fig:discrete_phylo_illust}.
